Manuscript doc

### 2021-09-21 paper/manuscript_word.Rmd
First commit, PNAS version

---
title: "Evidence of widespread maintenance of functional composition in vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Perry de Valpine"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: #word_document
  rticles::elsevier_article:
  includes:
   in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems [e.g. coral bleaching; @sully2019]. At the local scale however, species diversity tells a different story. Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017] .
<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species--> While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.

@@

While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001]. Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020]. And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020]. On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@barnosky2004; @jackson2015] .
It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for environmental impacts and management action.

@@

We obtained mammal, bird, and amphibian timeseries from the BioTIME database, a global repository of high quality assemblage timeseries.
All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018]. Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual timeseries following a standardized spatial scale. Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018] for details on how sample extents were defined), with the area of each cell set to one standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a single timeseries, and species abundances were combined for all samples.

```{r echo = FALSE, results = "asis"}
tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
       Timeseries_Count = c(48, 2380, 11),
       Species_Count = c(184, 700, 184),
       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
) %>%
  mutate_all(linebreak, align = "l") %>% 
  knitr::kable(format = "latex", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
               caption = "Summary of the data in the final trait database.") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
   #kableExtra::column_spec(2, width = "5cm") #%>%
  # kableExtra::column_spec(4, width = "10cm")
```

We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
These databases include species-level means for traits that best define species' function in the community including body size, diet, and behavioral characteristics.
For the full list of traits included in the analysis for each taxon see Table 1.
<!-- limitations of species level trait data? -->

In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (citation), obtained through the taxadb R package [@norman2020].
If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
For trait data, traits for all species of the same identifier were averaged.
Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.

@@

Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010]. Here we report functional richness (FRic), functional evenness (FEve), and functional divergence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
FRic assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
FEve describes how species are distributed across the trait space and how abundance is distributed across species.
Higher values of FEve indicate more even spacing of species in the trait space and individuals across species.
FDiv measures the degree to which species and their abundances maximize differences in the functional space.
Higher values of FDiv therefore correspond to communities where many highly abundant species are on the edges of the trait space.
<!-- use these metrics because they're common, not because of their results -->

@@

All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
All continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= s^t$, where $s$ is the number of species and *t* is the number of traits.
This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility.
Metrics incorporated weighting based on species abundance where available, three studies had data on presence only.

@@

We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each timeseries nested within the study.
To test for differences in trends by taxa, biome, and realm, each variable was modeled individually as random effects in models of the same structure.
We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model with no additional covariates.

We assessed the impact of timeseries duration and start year on study-level trends using general linear models with duration and start year as predictors.

@@

Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
However, a number of gaps in best practices may be obscuring a true underlying trend.
First, the BioTIME database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias. We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021]. We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.

@@

Third, while we use here the most common metrics for describing functional diversity they do not measure some potentially important aspects of the functional space.
Most notably, the metrics we calculated do not capture shifts in the location of the functional space.
For example, two communities could have very similar metric values but no overlap in their trait spaces.
This is especially relevant in the context of biodiversity change as a species loss could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
This scenario may be common in communities tracking changing environmental conditions.
Approaches for assessing shifts and overlaps in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.

@@

Our results illustrate a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological artifacts.
We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.
The most pressing need is for intentional and targeted data collection efforts.
We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017]. Data that fill geographic, taxonomic and trait gaps should be prioritized over further collection of data that replicate existing biases. One relatively low-cost high-reward data investment is collation of additional species-level trait means. Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases. Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks for understanding functional change [@lavorel2002].

>>>>

---
title: "Manuscript doc"
output: word_document
bibliography: refs.bib
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems (e.g. coral bleaching). At the local scale however, species diversity tells a different story. Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017] .
<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species--> While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.

@@

While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001]. Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020]. And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020]. On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@barnosky2004 @jackson2015] .
It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for environmental impacts and management action.

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime database, a global repository of high quality community assemblage timeseries.
All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018]. Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual timeseries following a standardized spatial scale. Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018] for details on how sample extents were defined), with the area of each cell set to one standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a single timeseries, and species abundances were combined for all samples.

```{r echo = FALSE, results = "asis"}
tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
       Timeseries_Count = c(48, 2380, 11),
       Species_Count = c(184, 700, 184),
       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
) %>%
  mutate_all(linebreak, align = "l") %>% 
  knitr::kable(format = "latex", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
               caption = "Summary of the data in the final trait database.") %>%
  kable_classic() #%>% 
  #kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
   #kableExtra::column_spec(2, width = "5cm") #%>%
  # kableExtra::column_spec(4, width = "10cm")
```

We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
These databases include species-level means for traits that best define species' function in the community including body size, diet, and behavioral characteristics.
For the full list of traits included in the analysis for each taxa see Table 1.
<!-- limitations of species level trait data? -->

In order to ensure taxonomic consistency across datasets, Biotime species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (citation), obtained through the taxadb R package [@norman2020].
If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
For trait data, traits for all species of the same identifier were averaged.
Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.

@@

Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010]. Here we report functional richness (FRic), functional evenness (FEve), and functional diveregence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
FRic assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
FEve describes how species are distributed across the trait space and how abundance is distributed across species.
Higher values of FEve indicate more even spacing of species in the trait space and individuals across species.
FDiv measures the degree to which species and their abundances maximize differences in the functional space.
Higher values of FDiv therefore correspond to communities where many highly abundant species are on the edges of the trait space.
<!-- use these metrics because they're common, not because of their results -->

@@

All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
All continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= s^t$, where $s$ is the number of species and *t* is the number of traits.
This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility.
Metrics incorporated weighting based on species abundance where available.

@@

We estimated general trends for each diversity metric using a Linear Mixed Effects model with a random slope and intercept for each timeseries nested within the study.
To test for differences in trends by taxa, biome, and realm, each variable was modeled individually as random effects in models of the same structure.
We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model with no additional covariates.

We assessed the impact of timeseries duration and start year on study-level trends using General Linear Models with duration and start year as predictors.

@@

Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
However, a number of gaps in best practices may be obscuring a true underlying trend.
First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias. We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021]. We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.

@@

Third, while we use here the most common metrics for describing functional diversity they do not measure some potentially important aspects of the functional space.
Most notably, the metrics we calculated do not capture shifts in the location of the functional space.
For example, two communities could have very similar metric values but no overlap in their trait spaces.
This is especially relevant in the context of biodiversity change as a species loss could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
This scenario may be common in communities tracking changing environmental conditions.
Approaches for assessing shifts and overlaps in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] a could shed critical insight into functional composition changes of this nature.

@@

Our results illustrates a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological artifacts.
We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.
The most pressing need is for intentional and targeted data collection efforts.
We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017]. Data that fill geographic, taxonomic and trait gaps should be prioritized over further collection of data that replicate existing biases. One relatively low-cost high-reward data investment is collation of additional species-level trait means. Intentional trait selection is critical for linking functional patterns to ecological process [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases. Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks for understanding functional change [@lavorel2002].
